---
title: "Multiplexed image analysis"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(timeout=10000)
```

This script contains the example analysis presented during the [Tumors and the immune system 2022 - High dimensional spatial profiling of tumor microenvironment](https://www.cb.uzh.ch/en/Education/Compulsory-courses/ModuleB.html) course.
It focuses on highly multiplexed image analysis after image processing and 
is a shortened version of the [IMC data analysis](https://github.com/BodenmillerGroup/IMCDataAnalysis)
book.

## Getting started

To follow the tutorial, please clone the repository:

```
git clone https://github.com/nilseling/MultiplexedImageAnalysis2022.git
```

or download the zipped version by clicking on `Code` and then `Download ZIP`.

The `MultiplexedImageAnalysis.Rmd` file contains runnable code to follow the tutorial.

## Install required packages

Please install [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/download/).

To follow the data analysis examples in this tutorial, you will need to 
install the following R packages from within R:

```{r, eval=FALSE}
install.packages("BiocManager")
BiocManager::install(c("cytomapper", "imcRtools", "openxlsx", "stringr", 
                       "dittoSeq", "tidyverse", "bluster", "patchwork",
                       "viridis", "scater", "scuttle", "S4Vectors", "igraph"))
```

Please also install [FIJI](https://imagej.net/software/fiji/) and [QuPATH](https://qupath.github.io/).

## Download example data {#download-data}

To highlight the basic steps of multiplexed image analysis, we provide example data that
were acquired as part of the **I**ntegrated i**MMU**noprofiling of large adaptive
**CAN**cer patient cohorts projects ([immucan.eu](https://immucan.eu/)). The
raw data of 4 patients can be accessed online at 
[zenodo.org/record/5949116](https://zenodo.org/record/5949116).
Image processing and segmentation has been performed using `steinbock` and we will only need to access the
sample/patient information here:

```{r download-sample-data}
download.file("https://zenodo.org/record/5949116/files/sample_metadata.xlsx", 
         destfile = "data/sample_metadata.xlsx")
```

### Processed multiplexed imaging data

The IMC raw data was processed using the 
[steinbock](https://github.com/BodenmillerGroup/steinbock) framework.
Image processing included file type conversion, cell segmentation and feature
extraction. 

**steinbock output**

The needed output of the `steinbock` framework includes the single-cell mean
intensity files, the single-cell morphological features and spatial locations,
spatial object graphs in form of edge lists indication cells in close proximity,
hot pixel filtered multi-channel images, segmentation masks, image metadata and
channel metadata. All these files will be downloaded here for easy access. The
commands which were used to process the data can be found in
`data/steinbock/steinbock.sh`.

```{r steinbock-results, warning=FALSE}
# download intensities
url <- "https://zenodo.org/record/6043600/files/intensities.zip"
destfile <- "data/steinbock/intensities.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock", overwrite=TRUE)
unlink(destfile)

# download regionprops
url <- "https://zenodo.org/record/6043600/files/regionprops.zip"
destfile <- "data/steinbock/regionprops.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock", overwrite=TRUE)
unlink(destfile)


# download neighbors
url <- "https://zenodo.org/record/6043600/files/neighbors.zip"
destfile <- "data/steinbock/neighbors.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock", overwrite=TRUE)
unlink(destfile)

# download images
url <- "https://zenodo.org/record/6043600/files/img.zip"
destfile <- "data/steinbock/img.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock", overwrite=TRUE)
unlink(destfile)

# download masks
url <- "https://zenodo.org/record/6043600/files/masks_deepcell.zip"
destfile <- "data/steinbock/masks_deepcell.zip"
download.file(url, destfile)
unzip(destfile, exdir="data/steinbock", overwrite=TRUE)
unlink(destfile)

# download individual files
download.file("https://zenodo.org/record/6043600/files/panel.csv", 
              "data/steinbock/panel.csv")
download.file("https://zenodo.org/record/6043600/files/images.csv", 
              "data/steinbock/images.csv")
download.file("https://zenodo.org/record/6043600/files/steinbock.sh", 
              "data/steinbock/steinbock.sh")
```

## Visualization of multiplexed images

The first part of the tutorial will present [FIJI](https://imagej.net/software/fiji/) and [QuPATH](https://qupath.github.io/) for interactive multiplexed image visualization.
Please make sure you have this software installed.

## Handling image and single-cell data in R

This section describes how to read in single-cell data and images into `R`
**after** image processing and segmentation (see Section \@ref(processing)).

To highlight examples for IMC data analysis, we provide already processed data at
[https://zenodo.org/record/6043600](https://zenodo.org/record/6043600).
This data has already been downloaded in Section \@ref(download-data) and can 
be accessed in the folder `data`.

We use the [imcRtools](https://github.com/BodenmillerGroup/imcRtools) package to
read in single-cell data extracted using the `steinbock` framework or the IMC
Segmentation Pipeline. Both image processing approaches also generate
multi-channel images and segmentation masks that can be read into `R` using the
[cytomapper](https://github.com/BodenmillerGroup/cytomapper) package.

```{r, message=FALSE}
library(imcRtools)
library(cytomapper)
```

### Reading in single-cell data

For single-cell data analysis in `R` the
[SingleCellExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html)
[@Amezquita2019] data container is commonly used within the Bioconductor
framework. It allows standardized access to (i) expression data, (ii) cellular
metadata (e.g. cell type), (iii) feature metadata (e.g. marker name) and (iv)
experiment-wide metadata. For an in-depth introduction to the `SingleCellExperiment`
container, please refer to the [SingleCellExperiment class](https://bioconductor.org/books/3.14/OSCA.intro/the-singlecellexperiment-class.html).

The [SpatialExperiment](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html)
class [@Righelli2021] is an extension of the `SingleCellExperiment` class. It
was developed to store spatial data in addition to single-cell data and an
extended introduction is accessible
[here](https://bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html).

To read in single-cell data generated by the `steinbock` framework or the IMC
Segmentation Pipeline, the `imcRtools` package provides the `read_steinbock` and
`read_cpout` function, respectively. By default, the data is read into a
`SpatialExperiment` object; however, data can be read in as a
`SingleCellExperiment` object by setting `return_as = "sce"`. All functions
presented in this book are applicable to both data containers.

The [steinbock](https://github.com/BodenmillerGroup/steinbock) framework was
used to process and segment IMC raw data available
[here](https://zenodo.org/record/5949116). In Section \@ref(download-data), the
processed data was downloaded from [here](https://zenodo.org/record/6043600) and
moved to the `data/steinbock` folder.

The `read_steinbock` function provided by `imcRtools` can now be used to read in
`steinbock` generated data. For more information, please refer to
`?read_steinbock`.

```{r read-steinbock}
spe <- read_steinbock("data/steinbock/")
spe
```

By default, single-cell data is read in as `SpatialExperiment` object. 
The summarized pixel intensities per channel and cell (here mean intensity) are
stored in the `counts` slot. Columns represent cells and rows represent channels.

```{r counts}
counts(spe)[1:5,1:5]
```

Metadata associated to individual cells are stored in the `colData` slot. After
initial image processing, these metadata include the numeric identifier (`ObjectNumber`),
the area, and morphological features of each cell. In addition, `sample_id` stores
the image name from which each cell was extracted and the width and height of the
corresponding images are stored.

```{r colData}
head(colData(spe))
```

The main difference between the `SpatialExperiment` and the
`SingleCellExperiment` data container in the current setting is the way spatial
locations of all cells are stored. For the `SingleCellExperiment` container, the
locations are stored in the `colData` slot while the `SpatialExperiment`
container stores them in the `spatialCoords` slot:

```{r spatialCoords}
head(spatialCoords(spe))
```

The _spatial object graphs_ generated by steinbock (see Section
\@ref(feature-extraction) are read into a `colPair` slot of the
`SpatialExperiment` (or `SingleCellExperiment`) object. Cell-cell interactions
(cells in close spatial proximity) are represented as "edge list" (stored as
`SelfHits` object). Here, the left side represents the column indices of the
"from" cells and the right side represents the column indices of the "to"
cells. For visualization of the _spatial object graphs_, please refer to
Section \@ref(spatial-viz).

```{r colPair}
colPair(spe, "neighborhood")
```

Finally, metadata regarding the channels are stored in the `rowData` slot. This
information is extracted from the `panel.csv` file. Channels are ordered by
isotope mass and therefore match the channel order of the multi-channel images
(see Section \@ref(read-images)).

```{r rowData}
head(rowData(spe))
```

After reading in the single-cell data, few further processing steps need to be
taken.

**Add additional metadata**

We can set the `colnames` of the object to generate unique identifiers per cell:

```{r set-colnames}
colnames(spe) <- paste0(spe$sample_id, "_", spe$ObjectNumber)
```

It is also often the case that sample-specific metadata are available externally.
For the current data, we need to link the cancer type (also referred to as "Indication")
to each sample. This metadata is available as external excel file:

```{r add-indication}
library(openxlsx)
library(stringr)
meta <- read.xlsx("data/sample_metadata.xlsx")
spe$patient_id <- as.vector(str_extract_all(spe$sample_id, "Patient[1-4]", simplify = TRUE))
spe$ROI <- as.vector(str_extract_all(spe$sample_id, "00[1-8]", simplify = TRUE))
spe$indication <- meta$Indication[match(spe$patient_id, meta$Sample.ID)]
unique(spe$indication)
```

The selected patients were diagnosed with different cancer types:

* SCCHN - head and neck cancer  
* BCC - breast cancer  
* NSCLC - lung cancer  
* CRC - colorectal cancer

**Transform counts**

The distribution of expression counts across cells is often observed to be
skewed towards the right side meaning lots of cells display low counts and few
cells have high counts. To avoid analysis biases from these high-expressing
cells, the expression counts are commonly transformed or clipped.

Here, we perform counts transformation using an inverse hyperbolic sine
function. This transformation is commonly applied to [flow cytometry
data](https://support.cytobank.org/hc/en-us/articles/206148057-About-the-Arcsinh-transform).
The `cofactor` here defines the expression range on which no scaling is
performed. While the `cofactor` for CyTOF data is often set to `5`, IMC data
usually display much lower counts. We therefore apply a `cofactor` of `1`.

However, other transformations such as `log(counts(spe) + 0.01)` should be
tested when analysing IMC data.

```{r transform-counts, message=FALSE}
library(dittoSeq)
dittoRidgePlot(spe, var = "CD3", group.by = "patient_id", assay = "counts") +
    ggtitle("CD3 - before transformation")
assay(spe, "exprs") <- asinh(counts(spe)/1)
dittoRidgePlot(spe, var = "CD3", group.by = "patient_id", assay = "exprs") +
    ggtitle("CD3 - after transformation")
```

**Define interesting channels**

For downstream analysis such as visualization, dimensionality reduction and
clustering, only a subset of markers should be used. As convenience,
we can store an additional entry in the `rowData` slot that specifies the 
markers of interest. Here, we deselect the nuclear markers and keep all other
biological targets.

```{r select-features}
rowData(spe)$use_channel <- !grepl("DNA|Histone", rownames(spe))
```

### Reading in images

The `cytomapper` package allows multi-channel image handling and visualization
within the Bioconductor framework. The most common data format for multi-channel
images or segmentation masks is the TIFF file format, which is used by `steinbock`
to save images. 

Here, we will read in multi-channel images and segmentation masks into a
[CytoImageList](https://www.bioconductor.org/packages/release/bioc/vignettes/cytomapper/inst/doc/cytomapper.html#5_The_CytoImageList_object)
data container. It allows storing multiple multi-channel images and requires
matched channels across all images within the object.

The `loadImages` function is used to read in processed multi-channel images and
their corresponding segmentation masks as generated by `steinbock`. The
multi-channel images are saved as 32-bit images while the segmentation masks are
saved as 16-bit images. To correctly scale pixel values of the segmentation
masks when reading them in set `as.is = TRUE`.

```{r read-images}
images <- loadImages("data/steinbock/img/")
masks <- loadImages("data/steinbock/masks_deepcell/", as.is = TRUE)
```

In the case of multi-channel images, it is beneficial to set the `channelNames`
for easy visualization. Using the `steinbock` framework, the channel order of the
single-cell data matches the channel order of the multi-channel images. However,
it is recommended to make sure that the channel order is known.

```{r set-channelNames}
channelNames(images) <- rownames(spe)
images
```

For visualization shown in Section \@ref(image-visualization) we will need to
add additional metadata to the `elementMetadata` slot of the `CytoImageList`
objects. This slot is easily accessible using the `mcols` function.

Here, we will save the matched `sample_id`, `patient_id` and `indication`
information within the `elementMetadata` slot of the multi-channel images and
segmentation masks objects. It is crucial that the order of the images in 
both `CytoImageList` objects is the same.

```{r add-metadata}
all.equal(names(images), names(masks))
patient_id <- str_extract_all(names(images), "Patient[1-4]", simplify = TRUE)
indication <- meta$Indication[match(patient_id, meta$Sample.ID)] 
mcols(images) <- mcols(masks) <- DataFrame(sample_id = names(images),
                                           patient_id = patient_id,
                                           indication = indication)
```

## Single-cell analysis

The single-cell data contained in the `SpatialExperiment` object can now be
analysed to extract biological knowledge. As part of the first exploratory analysis steps,
we want to visualize the cell density per mm$^2$. 

```{r density-cells-per-image, message=FALSE}
library(tidyverse)

colData(spe) %>%
    as.data.frame() %>%
    group_by(sample_id) %>%
    summarize(cell_count = n(),
           no_pixels = mean(width_px) * mean(height_px)) %>%
    mutate(cells_per_mm2 = cell_count/(no_pixels/1000000)) %>%
    ggplot() +
        geom_point(aes(sample_id, cells_per_mm2)) + 
        theme_minimal(base_size = 15)  + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("Cells per mm2") + xlab("")
```

A convenient way to visualize the protein expression in individual cells is using
heatmaps. In the following code chunk, we will randomly select 2000 cells
and visualize their expression.

```{r segmentation-heatmap, message=FALSE, fig.height=7}
library(dittoSeq)
library(viridis)
cur_cells <- sample(seq_len(ncol(spe)), 2000)
dittoHeatmap(spe[,cur_cells], genes = rownames(spe)[rowData(spe)$use_channel],
             assay = "exprs", cluster_cols = TRUE, scale = "none",
             heatmap.colors = viridis(100), annot.by = "indication")
```

An alternative way of visualizing high-dimensional data is to first perform
dimwnsionality reduction followed by scatter plot visualization of the 
cells in 2 dimensions. A common approach to perform non-linear dimensionality
reduction is Uniform Manifold Approximation and Projection (UMAP). The
[scater](https://bioconductor.org/packages/release/bioc/html/scater.html) Bioconductor package can be used to perform this dimensionality reduction.

```{r dimred, message=FALSE}
library(scater)
set.seed(220404)
spe <- runUMAP(spe, subset_row = rowData(spe)$use_channel, exprs_values = "exprs") 
```

After dimensionality reduction, the low-dimensional embeddings are stored in the 
`reducedDim` slot.

```{r show-dimred-slot}
head(reducedDim(spe, "UMAP"))
```

We can now visualize cells in low dimensions, color them by patient id
and visualize protein expression.

```{r visualizing-dimred, message=FALSE}
library(patchwork)

# visualize patient id 
p1 <- dittoDimPlot(spe, var = "patient_id", reduction.use = "UMAP", size = 0.2) +
    ggtitle("Patient ID on UMAP")

# visualize region of interest id
p2 <- dittoDimPlot(spe, var = "ROI", reduction.use = "UMAP", size = 0.2) +
    ggtitle("ROI ID on UMAP")

p1 + p2

# visualize marker expression
p1 <- dittoDimPlot(spe, var = "Ecad", reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2) +
    scale_color_viridis(name = "Ecad") +
    ggtitle("E-Cadherin expression on UMAP")
p2 <- dittoDimPlot(spe, var = "CD45RO", reduction.use = "UMAP", 
                   assay = "exprs", size = 0.2) +
    scale_color_viridis(name = "CD45RO") +
    ggtitle("CD45RO expression on UMAP")

p1 + p2
```

A common step in single-cell data analysis is the identification of cellular
phenotypes. This is often done by clustering cells based on their similarity
in protein expression. For clustering, we will use the [bluster](https://www.bioconductor.org/packages/release/bioc/html/bluster.html) package
and specifically using a graph-based clustering approach. 

```{r cluster, message=FALSE}
library(bluster)
mat <- t(assay(spe, "exprs")[rowData(spe)$use_channel,])

cluster <- clusterRows(mat, NNGraphParam(k=30, cluster.fun = "louvain"))
length(unique(cluster))

spe$cluster <- as.factor(cluster)
```

Using this approach, we detect 13 clusters. To annotate clusters, we will need
to visualize the protein expression per cluster. For this, we will visualize a
heatmap of the 2000 samples cells and observe single-cell cluster expression.

Based on the expression, the individual clusters can then be annotated.

```{r cluster-annotation}
dittoHeatmap(spe[,cur_cells], genes = rownames(spe)[rowData(spe)$use_channel],
             assay = "exprs", cluster_cols = FALSE, annot.by = "cluster",
             scaled.to.max = TRUE, heatmap.colors.max.scaled = viridis(100))

cluster_annot <- as.numeric(cluster)
cluster_annot[cluster_annot == 1] <- "Tumor_proliferating"
cluster_annot[cluster_annot == 2] <- "Stroma"
cluster_annot[cluster_annot == 3] <- "Tumor_proliferating"
cluster_annot[cluster_annot == 4] <- "Macrophages"
cluster_annot[cluster_annot == 5] <- "CD38"
cluster_annot[cluster_annot == 6] <- "Macrophages"
cluster_annot[cluster_annot == 7] <- "Tumor_hypoxic"
cluster_annot[cluster_annot == 8] <- "helper_Tcell"
cluster_annot[cluster_annot == 9] <- "cyto_Tcell"
cluster_annot[cluster_annot == 10] <- "Stroma"
cluster_annot[cluster_annot == 11] <- "Tumor"
cluster_annot[cluster_annot == 12] <- "Tumor_proliferating"
cluster_annot[cluster_annot == 13] <- "B cells"

spe$cluster_annot <- as.factor(cluster_annot)
```

We can now also visualize the location and interactions between the identified 
cell types in the tissue. For this, we select randomly 4 images.

```{r, fig.width=15,fig.height=15}
set.seed(1234)
cur_img_ids <- sample(unique(spe$sample_id), 4)

plotSpatial(spe[,spe$sample_id %in% cur_img_ids], img_id = "sample_id", 
            node_color_by = "cluster_annot", draw_edges = TRUE, colPairName = "neighborhood",
            nodes_first = FALSE)
```

This concludes the single-cell analysis and visulization and we will now move on
to the image-level visualization.

## Image visualization



```{r overlay-masks, message=FALSE}
library(cytomapper)
set.seed(20220118)
img_ids <- sample(seq_len(length(images)), 3)

# Normalize and clip images
cur_images <- images[img_ids]
cur_images <- cytomapper::normalize(cur_images, separateImages = TRUE)
cur_images <- cytomapper::normalize(cur_images, inputRange = c(0, 0.2))

plotPixels(cur_images,
           mask = masks[img_ids],
           img_id = "sample_id",
           missing_colour = "white",
           colour_by = c("CD163", "CD20", "CD3", "Ecad", "DNA1"),
           colour = list(CD163 = c("black", "yellow"),
                         CD20 = c("black", "red"),
                         CD3 = c("black", "green"),
                         Ecad = c("black", "cyan"),
                         DNA1 = c("black", "blue")),
           image_title = NULL,
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
```

## Software used

The following chunk lists all software packages used in this tutorial.

```{r SessionInfo}
sessionInfo()
```
